Application-Specific Enhanced Morphological Analyzer \\for Arabic Text Mining

### 2011-02-24 tex/sarf.aclhlt.tex
contains almost everything that needed to be added, but is about 1.5 pages off the limit till now

%\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}

%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
\title{Application Specific Arabic Morphological Analyzer for Arabic Text Mining}

@@

\begin{document}
\maketitle
\begin{abstract}
Natural language processing (NLP) uses morphological 
analysis to abstract and annotate text.
Often times, 
%      and in particular in the context of the Arabic language, 
annotations resulting from 
a branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, an application-specific morphological analyzer for Arabic. 
Sarf allows an application-specific controller to control and refine
the state of the analysis on the fly. 
It uses non-deterministic finite state Automata
to analyze a stream of text. 
%Each machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
We automated the analysis of three books of Islamic
narrations (hadith) using Sarf, where we
abstracted each book into a vector of complex structures.
Our results show higher accuracy and better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

%Automated analysis of Arabic data sets, including texts, 
%publications, records and digital media is essential
%with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
%Current morphological analyzers~\cite{Sughaiyer:04}
%use concatenative analysis when
%considering the internal structure 
%of an Arabic word and
%composing it into several {\em morphemes}. 
%A morpheme can be a {\em stem}, or an {\em affix}.
%An affix can be a {\em prefix, suffix, } or an {\em infix}.
%The analysis of one word may lead to several possible
%morphological solutions.
%\vocalize
%For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%may have two valid morphological analyses. 
%The letter \RL{'a} may be a prefix and the word means 
%``I praise him'', or 
%%The letter \RL{'a} may also be 
%part of the stem \RL{'a.hmad} (a proper noun)
%and the word means ``his Ahmad''.
%
%\novocalize
The accuracy of the solutions suffer due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels / diacritics. 
This greatly increases the ambiguity of Arabic text.
%(maybe reduce the couple of sentences below)
Moreover, Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase \transfalse
\RL{il_A\nospace almdrsT} \transtrue
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
without the need of a space in between. 
The reason is the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words,
referred to as ``run-on'' words~\cite{Buckwalter:04},
occur often, and greatly increase the
difficulty of tokenization.

%{\bf Sarf.~~~}
In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
has been submitted with the paper and will be available online as an open source tool.},
 a {\em novel application-specific efficient
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by an application-specific controller.
Each machine in Sarf takes one letter at a time as input
and represents a valid morphological analysis.
The application-specific controller prunes false positives
early in the run by making a decision at every input letter
to control the machines.
The controller we present in Section~\ref{sec:controller}
is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
it can be a logical entity of any type.

%Each alive machine in Sarf takes as input one letter at a time 
%from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

@@

%Our main focus when building Sarf, was not on improving the semantical rules and expanding the morphological alternatives, instead we worked on adding support for higher level control by the specific application utilizing the lower level morphological analyzer. In other words, the application-specific controller, has full control over and directs the morphological analysis of Sarf. % (sentence should be reworded differently, but is needed to explain why we mention higher and lower accuracy hypothesis, which seems vague otherwise)
%
%We hypothesize that many NLP case studies need the 
%morphological analyzer to answer queries that do not need 
%high accuracy at the low morphological analysis level.
%The analysis at the higher levels 
%(i.e. answering the application-specific query)
%can often compensate for 
%tolerable inaccuracy at lower morphological levels. 
%For example, if a query concerns proper names and the 
%analyzer is considering a 
%prefix that connects only to a verb,
%%the analyzer can answer without considering the 
%%rest of the word.
%we do not need to further analyze the rest of the word 
%to provide a negative answer.
%
%We find evidence to our hypothesis in~\cite{Maamouri:10}.
%%The addition of a new corpus to the Arabic Tree Bank 
%%(ATB)~\cite{Maamouri:04}
%%%required compliance with the DARPA GALE program 
%%%including several NLP tasks such as English translation 
%%%annotations and word-level alignment of Arabic and English. 
%%%The new data 
%%challenged the existing processing and 
%%annotation guidelines. 
%%This led to refining 
%%SAMA~\cite{Kulick:10} with 
%%alternative guidelines. 
%%This refinement was achieved only after the NLP 
%%task interacted with the morphological analyzer,
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
%It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when NLP task interacted with the morphological analyzer.
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run.
%In addition, 
%the work of Habash~\shortcite{Habash:06} reports that different 
%types of morphological analyses behave differently on the same %case study. 
%data set.
%The analyzers may report different solutions or 
%may present the solutions in different orders. 
%Clearly, some of 
%the presented solutions and their order of presentation may not be 
%appropriate for the %case study.
%application needs, which is an issue that Sarf tackles. % (maybe reworded)
%Moreover, 
Our results strongly support our claim that
an application-specific morphological analyzer that allows a application-specific 
controller to intervene at every decision to guide, use, order, and 
refine the morphological analysis is of high utility.

We validate our hypothesis using a case study from the Islamic 
literature, introduced later in Section~\ref{sec:islamic}.

@@

\section{Related work }
\label{sec:related}

@@

Current morphological analyzers such as 
Buckwalter~\shortcite{Buckwalter:02},
Beesley~\shortcite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
%This approach has several problems. 
However, we note the following points that deserve consideration:
%(We need not to mention especially point 2 as a problem according to reviewer comments)
First, a white space delimited token may have 
more than one word.
%Second, the exhaustive enumeration may hurt performance and may
%not be necessary or appropriate
%in some case studies as noted in~\cite{Maamouri:10}. 
Second, in many applications, the exhaustive enumeration may not be needed or appropriate as noted in~\cite{Maamouri:10}; consequently, adding unnecessary running time.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
% (reviewer wants to distinguish MADA as a tool for morphological disambiguation instead of analysis)
use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of performance. % (we should stress that we mean performance in time and not accuracy although evident since some reviewers did not get)
Xerox rules can be compiled into specialized finite state
machine (FSM) based analyzers as described in~\cite{Beesley:03}.
However, the efficiency of the resulting analyzer depends on the
way the Xerox rules are written. 
This requires deep knowledge and insight from the user
in compilation techniques, context free grammars, 
and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA).

@@

Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
finite state Automata (FSA). 
It reports that the number of Automata generated by a compiler
for Xerox rules can not be controlled and also reports a difficulty 
to compose the FSAs into a single framework. 
Sarf however constructs an NDFSA from
the controller and the prefix, stem, and suffix deterministic 
transducers to solve the problem.
Sarf also differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller to modify their behavior. 
%With Sarf it is easy to control the FSA with a case-based
%controller. 
Beesley requires recompilation of the Xerox rules. % (don't we do also for any change in the controller code?)

Similar to The MADA+TOKAN toolkit, Sarf thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
case study to interfere in judging the analyses much earlier and
at a finer granularity level. 
In Sarf, there is no need for a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The case-based controller can build the trace 
%when needed.

@@

Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. 
%We find supporting evidence in this decision to our hypothesis
%about the utility of a case-based analyzer. 
The algorithmic changes in SAMA were
done manually and worked in integration with the ATB format. 
Sarf is an application-specific controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with any case study. 
%In Sarf, the analyzer remains general, and the 
%case-based controller specializes the analysis.

@@

Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
of the Buckwalter analyzer. 
Sarf also uses deterministic parsing with tries 
to implement the affix and stem transducers. 
We think that the inferential-realizational approach 
of ElixirFM
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the application-specific controller
since in many cases the abstraction needs only a partial 
linguistic model of the Arabic language. 
%We can compare the case-based controller in Sarf to a high level
%component in the ElixirFM domain specific language.

@@

\transfalse
\begin{figure}[tb]
\center{
\resizebox{\columnwidth}{!}
{ \input{figs/FSM.pdftex_t}}
\caption{Example affix and stem transducers.}
\label{f:example}
}
\end{figure}
% the FSMs for the three words
\transtrue

@@

The diagram in Figure~\ref{f:example}
illustrates the operation of $\Phi$, the Sarf FSA,
when parsing the
string \RL{wsyl`b-h-aa al-lA`bwn}.
%\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
\RL{n}%
\RL{w}
\RL{b}
\RL{`}
\RL{a} 
\RL{l}
\RL{l}
\RL{a} %
\RL{a}
\RL{h}
\RL{b}
\RL{`}
\RL{l}
\RL{y}
\RL{s}
\RL{w}
in separate form to ease following the example
}

@@

When there are two valid transitions such as \RL{w} 
and $\epsilon$ in the start case, 
$\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ represent a valid analysis so far. 
An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
In our example, if there were no stems that start 
with the letter \RL{w}, $\Psi$ will die. 
In reality, $\Psi$ will die when the input 
is at \RL{wsyl`}.

Note that the affix transducers allow recursive 
affixes. 
For example, \RL{w} will result in an accept state
that transitions to ${\cal S}$.
When \RL{s} follows, we move to another accept state in 
${\cal P}$ corresponding to the prefix \RL{ws}. 
The same applies to ${\cal X}$.

Lets consider $\Phi$ after it consumed \RL{wsy} 
and transitioned into the root node of ${\cal S}$.
Now $\Phi$ will transition with \RL{l`b} to reach an accept 
state.

Before moving with the letter \RL{b} to the accept state,
$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
with the prefix \RL{wsy}. 
Sarf keeps compatibility category values as part
of the accept states. 
%Sarf considers the value of the category as part of the state.
Thus each accept state in Figure~\ref{f:example} represents
more than one concrete state. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in ${\cal S}$, it 
spawns $\Xi$ which transitions to the root of ${\cal X}$. 
After it consumes \RL{h--} $\Phi$ dies since there is no
\noTrRL{h--}-edge from the current state.
$\Xi$ represents a valid analysis and proceeds.

@@

The case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an FSA that 
is not interesting. 
It may also decide to correct one FSA
based on the decision of the others. 
The controller uses the output of the transducers as input
to make its decision. 
It does not produce external output and therefore 
the three transducers with the 
controller compose an FSA.

@@

The stems share lots of substrings. We encoded them in
an efficient double array trie structure~\cite{Aoe:89}. 
The structure is deterministic and free of cycles where the 
accept states are the terminal nodes corresponding to the last 
letters in stems. 
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need
a number of hash lookups in the order of all possible partitions
of the string.

@@

@@

The algorithm \CodeIn{NDSarf} (detailed below)
takes as input a text string $L$ and a case-based controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space, a delimiter, 
or when the preceding letter %leading to the current state
was a non-connecting letter.

@@

\begin{table}[tb]
%\centering
%\resizebox{.9\columnwidth}{!}{
\begin{tabular} {p{6cm}}
\begin{Verbatim}[fontsize=\relsize{-2},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Controller $A$) 
  MachineVec $M$; -- collection of machines
  Machine $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite()) or 
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

@@

\section{Islamic literature case study}
\label{sec:islamic}

% TODO: check how to merge this part here
\transfalse
\begin{figure}[tb]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/exhadith.pdftex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
% the FSMs for the three words
\transtrue

>>>>

%\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}
\newcommand{\noTrnoVocRL}[1]{\transfalse\novocalize\noTrRL{#1}\vocalize\transtrue}  
\newcommand{\noVocRL}[1]{\transtrue\novocalize\RL{#1}\vocalize}

%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
%\title{Application-Specific Arabic Morphological Analyzer for Arabic Text Mining}% maybe add "with Enhanced Capabilities"
\title{Application-Specific Enhanced Morphological Analyzer \\for Arabic Text Mining}% maybe add "with Enhanced Capabilities"

@@

\begin{document}
\maketitle
\begin{abstract}
Natural language processing (NLP) uses morphological 
analysis to abstract and annotate text.
Often times, 
%      and in particular in the context of the Arabic language, 
annotations resulting from 
a branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, an application-specific morphological analyzer for Arabic. 
Sarf allows an application-specific controller to control and refine
the state of the analysis on the fly. 
It uses non-deterministic finite state Automata
to analyze a stream of text. 
%Each machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
%We automated the analysis of three books of Islamic
%narrations (hadith) using Sarf, where we
%abstracted each book into a vector of complex structures.
Our results show better efficiency
compared to current analyzers.
\end{quote}
\end{abstract}

@@

%Automated analysis of Arabic data sets, including texts, 
%publications, records and digital media is essential
%with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
%Current morphological analyzers~\cite{Sughaiyer:04}
%use concatenative analysis when
%considering the internal structure 
%of an Arabic word and
%composing it into several {\em morphemes}. 
%A morpheme can be a {\em stem}, or an {\em affix}.
%An affix can be a {\em prefix, suffix, } or an {\em infix}.
%The analysis of one word may lead to several possible
%morphological solutions.
%\vocalize
%For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%may have two valid morphological analyses. 
%The letter \RL{'a} may be a prefix and the word means 
%``I praise him'', or 
%%The letter \RL{'a} may also be 
%part of the stem \RL{'a.hmad} (a proper noun)
%and the word means ``his Ahmad''.
%
%\novocalize
The accuracy of the solutions suffer due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels / diacritics. 
This greatly increases the ambiguity of Arabic text.
%(maybe reduce the couple of sentences below)
Moreover, Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase 
\noTrnoVocRL{il_A\nospace almdrsT}
to be visually recognizable
as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school) 
without the need of a space in between. 
The reason is the first word \noVocRL{il_A} ends with
\noVocRL{_A} a non-connecting letter. 
These words,
referred to as ``run-on'' words~\cite{Buckwalter:04},
occur often, and greatly increase the
difficulty of tokenization.

%{\bf Sarf.~~~}
In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
has been submitted with the paper and will be available online as an open source tool.},
 a {\em novel application-specific %efficient 
enhanced %(instead of 'efficient' to engulf both efficiency and other refinements)
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by an application-specific controller.
Each machine in Sarf takes one letter at a time as input
and represents a valid morphological analysis.
The application-specific controller prunes false positives
early in the run by making a decision at every input letter
to control the machines.
%The controller we present in Section~\ref{sec:controller}
%is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
%it can be a logical entity of any type.
The controller can be an FSA or any other %logical entity 
programming entity of any type,
since Automatas may be a very complex solution for some simple applications.
  %(maybe re-word this sentence)

%Each alive machine in Sarf takes as input one letter at a time 
%from a text stream.
In addition, to being application-specific, Sarf has introduced 
many enhancements to the field of arabic morphological analysis. 
For example,
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

@@

In addition, Sarf supports disambiguation using diacritics when they are available.
This feature, to the best of our knowledge is the first attempt to benefit from 
diacritics instead of ignoring them. Prior to Sarf, the trend was to preprocess 
the text to remove any such `unwelcomed' letter. %Nonetheless, Sarf regards them as a richness to be utilized that should not be disregarded.

%Moreover, since Sarf performs tokenization on the fly, Sarf is not just capable of solving
%the problem of ``run-on'' words, but also of handling Multiword Expressions (MWE). 
Moreover, since Sarf does not perform white-space delimited tokenization, Sarf is capable of 
handling Multiword Expressions (MWEs).
MWEs are defined in~\cite{MWE}
``as idiosyncratic interpretations
that cross word boundaries (or spaces)''. 
%By this term, we refer to frequent cases when a group of words form together a single
%item with a semantical meaning. In this paper, we focus on the example of compound names such as 
%\noVocRL{`bid alkarym}. 
%In current analyzers, this phrase is analyzed word by word and in best cases is 
%considered to contain two names, instead of being understood to be a single but compound name.

Our focus when building Sarf, was not on improving the semantical rules and 
expanding the morphological alternatives, instead we worked on adding support 
for higher level control by the specific application utilizing the lower level 
morphological analyzer. %In other words, the application-specific controller, has 
% full control over and directs the morphological analysis of Sarf. 
% (sentence should be reworded differently, but is needed to explain why we mention 
% higher and lower accuracy hypothesis, which seems vague otherwise)
%
%We have noted that many NLP case studies need the 
%morphological analyzer to answer queries that do not need 
%high accuracy at the morphological analysis level.
%The analysis at the higher levels 
%(i.e. answering the application-specific query)
%can often compensate for 
%tolerable inaccuracy at lower morphological levels. 
%For example, if a query concerns proper names and the 
%analyzer is considering a 
%prefix that connects only to a verb,
%%the analyzer can answer without considering the 
%%rest of the word.
%we do not need to further analyze the rest of the word 
%to provide a negative answer.
%
%We find evidence to our hypothesis in~\cite{Maamouri:10}.
%%The addition of a new corpus to the Arabic Tree Bank 
%%(ATB)~\cite{Maamouri:04}
%%%required compliance with the DARPA GALE program 
%%%including several NLP tasks such as English translation 
%%%annotations and word-level alignment of Arabic and English. 
%%%The new data 
%%challenged the existing processing and 
%%annotation guidelines. 
%%This led to refining 
%%SAMA~\cite{Kulick:10} with 
%%alternative guidelines. 
%%This refinement was achieved only after the NLP 
%%task interacted with the morphological analyzer,
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
%It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when NLP task interacted with the morphological analyzer.
%%a flexibility our analyzer is designed to provide for the application-specific controller on the run.
%In addition, 
%the work of Habash~\shortcite{Habash:06} reports that different 
%types of morphological analyses behave differently on the same %case study. 
%data set.
%The analyzers may report different solutions or 
%may present the solutions in different orders. 
%Clearly, some of 
%the presented solutions and their order of presentation may not be 
%appropriate for the %case study.
%application needs, which is an issue that Sarf tackles. % (maybe reworded)
We strongly believe that by being
an application-specific morphological analyzer that allows an application-specific 
controller to intervene at every decision to guide, use, order, and 
refine the morphological analysis, Sarf is of high utility.

%We validate our hypothesis using a case study from the Islamic 
%literature, introduced later in Section~\ref{sec:results}.
In addition to all its enhanced analysis methods, Sarf still
outperformed current morphological analyzers in terms of running time.
These results will be presented in Section~\ref{sec:results}.

We make the following key contributions: 
\begin{enumerate}
\item {\bf application-specific analysis:}  We designed and implemented
a application-specific morphological analyzer where a controller
guides the analysis. 
%\item {\bf Exhaustive analysis:} Since we have an application-specific 
%controller that can decide on the fly whether an analysis is 
%accepted or not, we can afford to keep all valid analyses modulo
%those pruned by the case controller. 
%We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02} and define affixes
recursively. 
This allows Sarf to perform better and 
maintain compositional part of speech (POS) tags.
\item {\bf Diacritic-based Disambiguation:}
Diacritics are used to disambiguate morphological
alternatives using any existant diacritics.
\end{enumerate}

@@

\subsection{Related work }
\label{sec:related}

@@

Current morphological analyzers such as %(maybe disambiguators also added here according to reviewer comments)
Buckwalter~\shortcite{Buckwalter:02},
Beesley~\shortcite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
%This approach has several problems. 
However, we note the following points that deserve consideration:
%(We need not to mention especially point 2 as a "problem" according to reviewer comments)
First, a white space delimited token may have 
more than one word or just a part of a MWE.
%Second, the exhaustive enumeration may hurt performance and may
%not be necessary or appropriate
%in some case studies as noted in~\cite{Maamouri:10}. 
Second, in many applications, the exhaustive enumeration may not be needed or appropriate as noted in~\cite{Maamouri:10}; consequently, adding unnecessary running time.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
% (reviewer wants to distinguish MADA as a tool for morphological disambiguation instead of analysis)
use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of efficiency.
Xerox rules can be compiled into specialized finite state
machine (FSM) based analyzers as described in~\cite{Beesley:03}.
However, the efficiency of the resulting analyzer depends on the
way the Xerox rules are written. 
This requires deep knowledge and insight from the user
in compilation techniques, context free grammars, 
and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA).

@@

Similar to Sarf, both Beesley~\shortcite{Beesley:01} and Attia~\shortcite{Attia:05} use
finite state Automata (FSA). 
Beesley reports that the number of Automata generated by a compiler
for Xerox rules can not be controlled and also reports a difficulty 
to compose the FSAs into a single framework. 
Sarf however constructs an NDFSA from
the controller and the prefix, stem, and suffix deterministic 
transducers to solve the problem.
Sarf also differs in that it uses recursive affixes in the 
affix transducers, and in that the transducers consider
input from the controller to modify their behavior. 
%With Sarf it is easy to control the FSA with a application-specific
%controller. 
Beesley requires recompilation of the Xerox rules. 
Attia has researched extensively MWEs handling methods~\cite{Attia:06} and 
automated extending their dictionary coverage~\cite{Attia:10}. 
% (I am not sure how clear this statement is)

Similar to The MADA+TOKAN toolkit, Sarf thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
application to interfere in judging the analyses much earlier and
at a finer granularity level. 
In Sarf, there is no need for a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The application-specific controller can build the trace 
%when needed.

@@

Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. 
%We find supporting evidence in this decision to our hypothesis
%about the utility of a application-specific analyzer. 
The algorithmic changes in SAMA were
done manually and worked in integration with the ATB format. 
Sarf is an application-specific controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with any application as needed. 
In Sarf, the analyzer remains general, and the 
application-specific controller specializes the analysis.

@@

Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
of the Buckwalter analyzer. 
Sarf also uses deterministic parsing with tries 
to implement the affix and stem transducers. 
We think that the inferential-realizational approach 
of ElixirFM
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the application-specific controller
since in many cases the abstraction needs only a partial 
linguistic model of the Arabic language. 
%We can compare the application-specific controller in Sarf to a high level
%component in the ElixirFM domain specific language.

@@

\transfalse
\begin{figure}[tb]
\center{
\resizebox{\columnwidth}{!}
{ \input{figs/FSM.pdftex_t}}
\caption{Example affix and stem transducers.}
\label{f:example}
}
\end{figure}
% the FSMs for the three words
%\transtrue

@@

The diagram in Figure~\ref{f:example}
illustrates the operation of $\Phi$, the Sarf FSA,
when parsing the
string \noTrnoVocRL{wsyl`b-h-aa\nospace al-lA`bwn}.
%\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
\noVocRL{n}%
\noVocRL{w}
\noVocRL{b}
\noVocRL{`}
\noVocRL{a} 
\noVocRL{l}
\noVocRL{l}
\noVocRL{a} %
\noVocRL{a}
\noVocRL{h}
\noVocRL{b}
\noVocRL{`}
\noVocRL{l}
\noVocRL{y}
\noVocRL{s}
\noVocRL{w}
in separate form to ease following the example
}

@@

When there are two valid transitions such as \noTrnoVocRL{w} 
and $\epsilon$ in the start case, 
$\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ represent a valid analysis so far. 
An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
In our example, if there were no stems that start 
with the letter \noTrnoVocRL{w}, $\Psi$ will die. 
In reality, $\Psi$ will die when the input 
is at \noTrnoVocRL{wsyl`}.

Note that the affix transducers allow recursive 
affixes. 
For example, \noTrnoVocRL{w} will result in an accept state
that transitions to ${\cal S}$.
When \noTrnoVocRL{s} follows, we move to another accept state in 
${\cal P}$ corresponding to the prefix \noTrnoVocRL{ws}. 
The same applies to ${\cal X}$.

Lets consider $\Phi$ after it consumed \noTrnoVocRL{wsy} 
and transitioned into the root node of ${\cal S}$.
Now $\Phi$ will transition with \noTrnoVocRL{l`b} to reach an accept 
state.

Before moving with the letter \noTrnoVocRL{b} to the accept state,
$\Phi$ needs to make sure that the stem \noTrnoVocRL{l`b} is compatible
with the prefix \noTrnoVocRL{wsy}. 
Sarf keeps compatibility category values as part
of the accept states. 
%Sarf considers the value of the category as part of the state.
Thus each accept state in Figure~\ref{f:example} represents
more than one concrete state. 
If the category of \noTrnoVocRL{l`b} is compatible with the category of
\noTrnoVocRL{wsy} then \noTrnoVocRL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in ${\cal S}$, it 
spawns $\Xi$ which transitions to the root of ${\cal X}$. 
After it consumes \noTrnoVocRL{h--} $\Phi$ dies since there is no
\noTrnoVocRL{h--}-edge from the current state.
$\Xi$ represents a valid analysis and proceeds.

@@

The application-specific controller can interfere at any point in the 
analysis to make a decision like ignoring an FSA that 
is not interesting. 
It may also decide to correct one FSA
based on the decision of the others. 
The controller uses the output of the transducers as input
to make its decision. 
It does not produce external output and therefore 
the three transducers with the 
controller compose an FSA.

@@

The stems share lots of substrings. We encoded them in
an efficient double array trie structure~\cite{Aoe:89}. 
The structure is deterministic and free of cycles where the 
accept states are the terminal nodes corresponding to the last 
letters in stems. Note that the trie implementation may not 
necessary work on letter-by-letter states, but also compacts
common substrings.
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need
a number of hash lookups in the order of all possible partitions
of the string.

@@

%Note that Sarf does not rely on implementing its machines as 
%equivalent regular expressions due to  the following reasons ...
% (not necessary)

@@

The algorithm \CodeIn{NDSarf} (detailed below)
takes as input a text string $L$ and a application-specific controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:composition} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space, a delimiter, 
or when the preceding letter %leading to the current state
was a non-connecting letter.

@@

\begin{table}[tb]
%\centering
%\resizebox{.9\columnwidth}{!}{
\begin{tabular} {p{6cm}}
\begin{Verbatim}[fontsize=\relsize{-2},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Controller $A$) 
  MachineVec $M$; -- collection of machines
  Machine $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite())  
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

@@

\subsection{Disambiguation using diacritics}
\label{sec:diacritics}

Many Arabic documents, especially religious texts, are fully vocalized.
Even in documents written without diacritics, we often stumble at a couple of words which are partially vocalized
to reduce the difficulty of their disambiguation by the reader. Clearly, if native readers
need the aid of short vowels, automatic morphological analyzers which are context-free should not lose this precious piece of information.
Sarf is the first in the field to support diacritic-based disambiguation. Consequently, Sarf does not confuse \RL{bon} (coffee beans) with \RL{bin} (child)
or \RL{`ilm} (science) with \RL{`lim} (knew), regardless of the specific letters which are vocalized.

The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.

\subsection{Multiword Expressions}
\label{sec:MWE}
